Reasonably Programmable Literal Notation

### 2018-02-19 pldi18/syntax-pldi18.tex
forgot to commit these

@@

%% Listings
\definecolor{mygray}{rgb}{0.75,0.75,0.75}
\usepackage{listings}
\lstset{tabsize=2, 
basicstyle=\ttfamily\fontsize{8pt}{1em}\selectfont, 
commentstyle=\itshape\ttfamily\color{gray},
stringstyle=\ttfamily\color{red},
mathescape=false,escapechar=~,
numbers=left, numberstyle=\scriptsize\color{mygray}, language=ML,moredelim=[il][\sffamily]{?},showspaces=false,showstringspaces=false,xleftmargin=0pt, numbersep=-6pt, morekeywords=[1]{try,spliced,tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,string,datatype,new,toast,syntax,module,switch,where,using,import,for,ana,syn,opcon,tycon,metasignature,metamodule,metasig,metamod,static,at,by,tycase,mod,macro,match,pattern,in,patterns,expressions,implicit,forall,exptsm,pattsm},deletekeywords={double},classoffset=0, xleftmargin=-10pt, 
aboveskip=4pt,belowskip=2pt,
moredelim=**[is][\color{red}]{SSTR}{ESTR},
moredelim=**[is][\color{OliveGreen}]{SHTML}{EHTML},
moredelim=**[is][\color{purple}]{SCSS}{ECSS},
moredelim=**[is][\color{brown}]{SSQL}{ESQL},
moredelim=**[is][\color{orange}]{SCOLOR}{ECOLOR},
moredelim=**[is][\color{magenta}]{SPCT}{EPCT}, 
moredelim=**[is][\color{gray}]{SNAT}{ENAT}, 
moredelim=**[is][\color{OliveGreen}]{SURL}{EURL},
moredelim=**[is][\color{SeaGreen}]{SQT}{EQT},
moredelim=**[is][\color{Periwinkle}]{SGRM}{EGRM},
moredelim=**[is][\color{YellowGreen}]{SID}{EID},
moredelim=**[is][\color{Sepia}]{SUS}{EUS},
deletestring=[d]{"},
}
\lstset{morecomment=[n]{/*}{*/}}

@@

Compared to existing approaches, including direct grammar extension systems and prior  macro-based parsing systems, TLMs are uniquely \emph{reasonable}: TLM providers can reason modularly about syntactic ambiguity, and TLM clients can reason abstractly, i.e. without examining the underlying expansion, about types and binding. The system only needs to convey to clients via secondary notation the inferred \emph{segmentation} of each literal, which gives the locations of spliced subterms. 

@@

We are integrating TLMs into Reason, an alternative front-end for OCaml.% This calculus is the first such formalization of a typed macro system.

@@

% \vspace{-4px}
\newcommand{\introSec}{Introduction}
\section{\protect\introSec}
\label{sec:intro}
% The surface syntax of a programming language is its user interface. 
When designing the surface syntax of a general-purpose programming language, it is common practice to define \emph{literal notation} that decreases the syntactic cost of constructing and pattern matching over values of a few data structures. For example, in Standard ML (SML), the list literal \li{[x, y, z]} is equivalent by desugaring to \li{Cons(x,Cons(y,Cons(z,Nil)))} where \li{Nil} and \li{Cons} stand for the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}. 
%\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e. the meaning of literal forms is \emph{context independent}. 
% We will return to this concept again later.} 
Similarly, Ur/Web's surface syntax extends Ur's syntax with expression and pattern literals for encodings of XML and HTML data \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}. Figure \ref{fig:urweb} shows two such HTML literals, one that ``splices in'' a string expression (Line 1) and the other an HTML expression (Line 2).
%The desugarings of these literal expressions, not shown, are substantially more verbose and less readable to programmers familiar with XHTML.

@@

Although requests for specialized literal notation are easy to dismiss as superficial, the reality is that literal notation, or the absence thereof, can have a substantial influence on software quality. For example, \citet{Bravenboer:2007:PIA:1289971.1289975} finds that literal notation for structured encodings of queries, like the SQL-based query literals now found in many languages \cite{meijer2006linq}, reduce the temptation to use string encodings of queries and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on web servers today \cite{owasp2017}. More generally, evidence suggests that programmers frequently engage in ``stringly-typed programming'', i.e. they choose strings instead of composite data structures largely for reasons of notational convenience. In particular, \citet{TSLs} sampled strings from open source projects and found that about 15\% of them were parseable according to some readily apparent type-specific grammar, e.g. for SQL queries, regexes, URLs and many other data structures. Literal forms, with support for splicing, would decrease the syntactic cost of composite encodings of such data, which are more amenable to programmatic manipulation and compositional reasoning than flat string encodings. %Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.

\begin{figure*}
\vspace{-2px}
\begin{lstlisting}
  fun heading first_name = SURL<xml><h1>Hello, {[EURLfirst_nameSURL]}!</h1></xml>EURL
  val body = SURL<xml><body>{EURLheading "World"SURL} ...</body></xml>EURL
\end{lstlisting}
\vspace{-10px}
\caption{HTML literals with support for splicing at two different types are built primitively into Ur/Web \cite{conf/popl/Chlipala15}.}
\vspace{-10px}
\label{fig:urweb}
\end{figure*}

@@

\paragraph{Problem} The problem that motivates this paper is that it is impossible to reason abstractly, i.e. by inspecting only the surface syntax, about such fundamental issues as types and variable binding when the program text is littered with  unfamiliar user-defined literal forms. Instead, programmers can only reason transparently, i.e. by inspecting the underlying expansion or the extension responsible for producing the expansion. In some existing systems, even determining which extension is responsible for a literal form is difficult.
%, i.e. when developing large programs ``consisting of many small programs (modules), possibly written by different people''

\begin{figure}[t!]
\begin{lstlisting}
  let w = compute_w();
  let x = compute_x(w);
  let y = [|(!R)@&{&/x!/:2_!x}'!R}|];
\end{lstlisting}
\vspace{-10px}
\caption{Syntax extension systems produce unreasonable program text.}
\vspace{-11px}
\label{fig:K-dialect}
\end{figure}

Consider, for example, the perspective of a programmer attempting to comprehend, i.e. reason about, the program text in Figure \ref{fig:K-dialect}, which is written in a dialect of OCaml's syntax called Reason \cite{reason} that has, hypothetically, been extended with some large number of new literal forms by a grammar extension system, like Camlp4 \cite{ocaml-manual} or Sugar* \cite{erdweg2011sugarj,erdweg2013framework}. 
The literal form on Line 3 constructs an encoding of a query in the niche stack-based database query language K, using its intentionally terse notation \cite{Whitney:2001:LOR:376284.375783}. The problem is that a programmer unfamiliar with this notation cannot fall back on abstract reasoning principles to answer questions like:

\begin{enumerate}[leftmargin=10pt]
\item \textbf{(Responsibility)} Which syntax extension determined the expansion of the literal form on Line 3? Might different extensions generate different expansions for this form (i.e. is there syntactic ambiguity)?
\item \textbf{(Segmentation)} Are the characters \li{x} and \li{R} on Line 3 parsed as spliced expressions (meaning that they appear directly in the underlying expansion), or are they parsed in some other way peculiar to this literal form (e.g. as operators in the K query language)?
\item \textbf{(Capture)} If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 2, or might it capture an unseen binding of the same identifier in the expansion of Line 3?
\item \textbf{(Context Dependence)} If \li{w}, on Lines 1-2, is renamed, could that possibly break Line 3 or change its meaning because the expansion assumes that \li{w} is in scope?% What other hidden assumptions might the desugaring be making?
\item \textbf{(Typing)} What type does \li{y} have?
\end{enumerate}

% In short, the problem is that library providers cannot reason modularly about syntactic determinism (because other extensions might define overlapping forms), and if the desugaring of the program text is held abstract, programmers can no longer reason about types, binding and segmentation (i.e. answer questions like those above.) This is burdensome at all scales, but particularly when programming in the large, where it is common to encounter a wide variety of libraries \cite{DBLP:conf/sac/LammelPS11}. 
Forcing the programmer to reason transparently to answer basic questions like these defeats the ultimate purpose of syntactic sugar: decreasing cognitive cost \cite{Green89}. Analagous problems do not arise when programming without syntax extensions in languages like ML and Scala -- the binding discipline of the language ensures that programmers can reason lexically about binding, and types mediate abstraction over function and module implementations. % From this perspective, it is unsurprising that the use of Camlp4 has been deprecated in the OCaml system.

An approach that has made some headway against these problems is Omar et al.'s \emph{type-specific languages} (TSLs) \cite{TSLs}. In this approach, the language delegates control over the parsing and expansion of certain \emph{generalized literal forms} to a parse function associated with the type that the form is being checked against. Generalized literal forms syntactically subsume other literal forms because their bodies (i.e. the characters between the outer delimiters, here \li{[|} and \li{|]}) are left  unparsed by the fixed context-free syntax of the language (i.e. they are initially parsed like raw string literals.) With TSLs, Line 3 of our example would require a type annotation:
\begin{lstlisting}[numbers=none]
    let y : kquery = [|(!R)@&{&/x!/:2_!x}'!R}|]
\end{lstlisting}
The parse function associated with the \li{kquery} type would be invoked to parse and expand the literal body.

Although TSLs address the problems of \textbf{Responsibility} and \textbf{Typing}, there is no guarantee that a valid segmentation exists, nor was the problem of reasoning about binding (i.e. capture and context dependence) adequately addressed in the prior work. Moreover, this prior work was introduced in the context of a very simple monomorphic, nominally typed language. There was no support for important features of modern languages in the ML tradition, such as structural types (e.g. arrow and tuple types), parameterized types, abstract types, modules, pattern notation and ML-style type inference.  Finally, parse functions and the expansions they generate needed to be closed, i.e. they could not make use of any libraries at all. Taken together, we found that TSLs were not suitable as-is for integration into Reason, and other languages with a broadly similar type and binding discipline (e.g. Haskell, Scala, Rust and so on.)

\paragraph{Contributions} The aim of this paper is to address all of these deficiencies, resulting in the first system for defining new literal notation that (1) provably maintains all five of the abstract reasoning principles just outlined; and (2) that is expressive enough for integration into a full-scale typed functional language (in this paper, the Reason front-end for OCaml.) We evaluate the first claim by developing a series of formal calculi equipped with proofs of these reasoning principles, and the second claim with several non-trivial case studies that appear throughout the paper. The empirical evaluation in the prior work on TSLs \cite{TSLs} supports the claim that a mechanism for defining new literal notation is useful, so our focus is on formal reasoning principles and integration with advanced language features in this paper.

\begin{figure}
\begin{lstlisting}
  let w = compute_w();
  let x = compute_x(w);
  let y = $kquery [|SURL(!R)@&{&/EURLxSURL!/:2_!EURLxSURL}'!R}EURL|];
\end{lstlisting}
\vspace{-10px}
\caption{TLMs make examples like the one from Figure \ref{fig:K-dialect} more reasonable.}
\vspace{-10px}
\label{fig:K-tsm-example}
\end{figure}

The most obvious point of departure between TSLs and \emph{typed literal macros} (TLMs), which are the topic of this paper, is that we separate the mechanism for defining new literal notation from the type definition mechanism of the language. TLMs are explicitly named and applied. For example, Figure \ref{fig:K-tsm-example} applies a TLM named \li{$kquery} to a generalized literal form delimited by \li{[|} and \li{|]} to express the K query example just discussed. The applied TLM parses the literal body and generates an expansion statically.

During the typing phase, a validation step ensures that programmers can reason abstractly about the five issues just identified. We will characterize these reasoning principles precisely as we proceed, but for now, let us develop the basic intuition -- the programmer can reason abstractly as follows:
\begin{enumerate}[nolistsep,leftmargin=10pt]
\item \textbf{(Responsibility)} The applied TLM, \li{$kquery}, parses and determines the expansion of the literal body. The context-free syntax of the language is fixed, so the TLM provider is able to modularly establish that the syntax that this parser implements is unambiguous (even in situations where a spliced term itself applies a TLM, not shown here.)
\item \textbf{(Segmentation)} The intermediate output that the TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms from segments that are parsed in some other way by the TLM. The segmentation can be communicated by a code editor using secondary notation, e.g. colors in this document. So by examining Figure \ref{fig:K-tsm-example}, the programmer knows that the two instances of \li{x} on Line 3 are parsed as spliced expressions (because they are black), whereas the \li{R}'s must be parsed in some other way, e.g. as operators in the K language (because they are  green.)
\item \textbf{(Capture)} Splicing is capture-avoiding, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2. It cannot capture a coincidental binding of \li{x} in the expansion.
\item \textbf{(Context Dependence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, e.g., \li{w} is in scope. The client is free to rename \li{w}.
\item \textbf{(Typing)} A type annotation on the definition of \li{$kquery} (not shown) determines the type that the expansion must have. We will see an example of a TLM definition in the next section. Moreover, each spliced segment in the segmentation also comes with a type annotation. %This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired. %TLM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kquery}.
\end{enumerate}

@@

\begin{subfigure}[t]{\textwidth}
\begin{lstlisting}
  let heading first_name = $html [|SURL<h1>Hello, {[EURLfirst_nameSURL]}!</h1>EURL|]
  let body = $html [|SURL<body>{EURLheading "World!"SURL} ...</body>EURL|]
\end{lstlisting}
\vspace{-5px}
\caption{Two examples of the \li{$html} TLM being applied. Compare to Ur/Web's built in HTML literal notation from Figure \ref{fig:urweb}.  %In each case, literal body, between backticks, is initially left unparsed according to the language's context-free syntax. %The applied TLM determines a segmentation and expansion during the typed expansion phase, which generalizes the usual typing phase.
}
\label{fig:first-tsm-example}
\end{subfigure}
\vspace{-8px}
\caption{Case Study: HTML literals}
\vspace{-5px}
\end{figure*}

\paragraph{Outline} 
% \begin{itemize}
% \item 
We begin in Sec.~\ref{sec:setsms} with a more detailed case study: a TLM that implements the HTML literal notation shown in Figure \ref{fig:urweb}. We use this case study as a running example to informally describe how TLMs operate, from start to finish.
% \item 
Sec.~\ref{sec:sptsms} then introduces {pattern TLMs}, i.e. TLMs that generate patterns, rather than expressions. %The hygiene condition for pattern TLMs is interesting.
% \item 
Sec.~\ref{sec:ptsms} adds type functions and an ML-style module system and introduces \emph{parametric TLMs} (pTLMs), i.e. TLMs that can take type and module parameters. This section is organized around a second case study: finite map literals. 
Sec.~\ref{sec:static-eval} considers the topic of term evaluation during the  expansion phase, i.e. \emph{static evaluation}, in  more detail. Sec \ref{sec:static-eval} also gives examples of TLMs that are useful for defining other TLMs, e.g. TLMs that implement parser generators and quasiquotation, and uses these in a third case study of the mechanism: IPv4 address literals. 
Sec.~\ref{sec:setsms-formally} defines a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above. 
% \item 
Sec.~\ref{sec:existing-approaches} compares TLMs to related work using the rubric of five reasoning principles just discussed. 
% \item 
Sec.~\ref{sec:discussion} concludes with a summary of the specific contributions of this paper and a discussion of the limitations of TLMs, together with suggestions for future work.

@@

Due to the space limitations, certain technical details and extensions to the calculi described in this paper, as well as detailed proofs, are given in the supplemental material. Some notes on implementation and a fourth case study, involving an encoding of regular expressions as an abstract data type, are also both given in the supplement.

@@

\vspace{-4px}
\newcommand{\seTLMsSec}{Expression TLMs}
\section{\protect\seTLMsSec}
\label{sec:setsms}
\vspace{-2px}

Let us begin by detailing the TLM mechanism by way of a case study: HTML literal notation for expressions of the \li{html} type defined in Figure \ref{fig:html-type-def}.

\vspace{-4px}
\subsection{TLM Definition and Application}
\vspace{-2px}
The definition of the \li{$html} TLM is given in Figure \ref{fig:html-tlm-def}. TLM names are prefixed by \li{$} 
 to distinguish them from variables. Every TLM definition also provides a \emph{type annotation}, here \li{at html}, and a \emph{parse function} between \li{by static \{} and \li{\}}. % The TLM definition follows the same scoping rules as type definitions. We will describe how TLM definitions are packaged into libraries in Sec \ref{sec:static-eval}.

>>>>

% \usepackage{titlesec}

% \titlespacing*\section{0pt}{1pt plus 2pt minus 2pt}{2pt plus 2pt minus 2pt}
% \titlespacing*\subsection{0pt}{1pt plus 2pt minus 2pt}{2pt plus 2pt minus 2pt}
% \titlespacing*\subsubsection{0pt}{1pt plus 2pt minus 2pt}{2pt plus 2pt minus 2pt}
% \titlespacing*\paragraph{0pt}{5pt plus 2pt minus 2pt}{2pt plus 2pt minus 2pt}

@@

%% Listings
\definecolor{mygray}{rgb}{0.75,0.75,0.75}
\usepackage{listings}
\lstset{tabsize=2, 
basicstyle=\ttfamily\fontsize{8pt}{1em}\selectfont, 
commentstyle=\itshape\ttfamily\color{gray},
stringstyle=\ttfamily\color{red},
mathescape=false,escapechar=~,
numbers=left, numberstyle=\scriptsize\color{mygray}, language=ML,moredelim=[il][\sffamily]{?},showspaces=false,showstringspaces=false,xleftmargin=0pt, numbersep=-6pt, morekeywords=[1]{try,spliced,tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,string,datatype,new,toast,syntax,module,switch,where,using,import,for,ana,syn,opcon,tycon,metasignature,metamodule,metasig,metamod,static,at,by,tycase,mod,macro,match,pattern,in,patterns,expressions,implicit,forall,exptsm,pattsm},deletekeywords={double,structure,of},classoffset=0, xleftmargin=-10pt, 
aboveskip=4pt,belowskip=2pt,
moredelim=**[is][\color{red}]{SSTR}{ESTR},
moredelim=**[is][\color{OliveGreen}]{SHTML}{EHTML},
moredelim=**[is][\color{purple}]{SCSS}{ECSS},
moredelim=**[is][\color{brown}]{SSQL}{ESQL},
moredelim=**[is][\color{orange}]{SCOLOR}{ECOLOR},
moredelim=**[is][\color{magenta}]{SPCT}{EPCT}, 
moredelim=**[is][\color{gray}]{SNAT}{ENAT}, 
moredelim=**[is][\color{OliveGreen}]{SURL}{EURL},
moredelim=**[is][\color{SeaGreen}]{SQT}{EQT},
moredelim=**[is][\color{Periwinkle}]{SGRM}{EGRM},
moredelim=**[is][\color{YellowGreen}]{SID}{EID},
moredelim=**[is][\color{Sepia}]{SUS}{EUS},
deletestring=[d]{"},
}
\lstset{morecomment=[n]{/*}{*/}}

@@

Compared to existing approaches, including direct grammar extension systems and prior  macro-based parsing systems, TLMs are uniquely \emph{reasonable}: TLM providers can reason modularly about syntactic ambiguity, and TLM clients can reason abstractly, i.e. without examining the underlying expansion, about types and binding. The system only needs to convey to clients, via secondary notation, the inferred \emph{segmentation} of each literal, which gives the locations of spliced subterms. 

@@

We are integrating TLMs into Reason, an emerging alternative front-end for OCaml.% This calculus is the first such formalization of a typed macro system.

@@

\newcommand{\introSec}{Introduction}
\section{\protect\introSec}
\label{sec:intro}
\vspace{-3px}
% The surface syntax of a programming language is its user interface. 
When designing the surface syntax of a general-purpose programming language, it is common practice to define \emph{literal notation} that decreases the syntactic cost of constructing and pattern matching over values of a few data structures. For example, in Standard ML (SML), the list literal \li{[x, y, z]} is equivalent by desugaring to \li{Cons(x,Cons(y,Cons(z,Nil)))} where \li{Nil} and \li{Cons} stand for the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}. 
%\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e. the meaning of literal forms is \emph{context independent}. 
% We will return to this concept again later.} 
Similarly, Ur/Web's surface syntax extends Ur's syntax with expression and pattern literals for encodings of XML and HTML data \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}. Figure \ref{fig:urweb} shows two HTML literals, one that ``splices in'' a string expression (Line 1) and the other an HTML expression (Line 2).
%The desugarings of these literal expressions, not shown, are substantially more verbose and less readable to programmers familiar with XHTML.

@@

Although requests for specialized literal notation are easy to dismiss as superficial, the reality is that literal notation, or the absence thereof, can have a substantial influence on software quality. For example, \citet{Bravenboer:2007:PIA:1289971.1289975} finds that literal notation for structured encodings of queries, like the SQL-based query literals now found in many languages \cite{meijer2006linq}, reduce the temptation to use string encodings of queries and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on web servers today \cite{owasp2017}. More generally, evidence suggests that programmers frequently engage in ``stringly-typed programming'', i.e. they choose strings instead of composite data structures largely for reasons of notational convenience. In particular, \citet{TSLs} sampled strings from open source projects and found that about 15\% of them could be parsed by some readily apparent type-specific grammar. Literal notation, with support for splicing, would decrease the syntactic cost of composite encodings of such data, which are more amenable to programmatic manipulation and compositional reasoning than flat strings. %Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.

\begin{figure*}
\begin{lstlisting}
  fun heading first_name = SURL<xml><h1>Hello, {[EURLfirst_nameSURL]}!</h1></xml>EURL
  val body = SURL<xml><body>{EURLheading "World"SURL} ...</body></xml>EURL
\end{lstlisting}
\vspace{-10px}
\caption{HTML literals with support for splicing at two different types are built primitively into Ur/Web \cite{conf/popl/Chlipala15}.}
\vspace{-10px}
\label{fig:urweb}
\end{figure*}

@@

\paragraph{Problem} The problem that motivates this paper is that it is impossible to reason abstractly, i.e. by inspecting only the surface syntax, about such fundamental issues as types and variable binding when the program text is littered with  unfamiliar user-defined literal forms. Instead, programmers can only reason transparently, i.e. by inspecting the underlying expansion or the code responsible for producing the expansion. In some existing systems, even determining which extension is responsible for a literal form is difficult.
%, i.e. when developing large programs ``consisting of many small programs (modules), possibly written by different people''

\begin{figure}[t!]
\begin{lstlisting}
  let w = compute_w();
  let x = compute_x(w);
  let y = [|(!R)@&{&/x!/:2_!x}'!R}|];
\end{lstlisting}
\vspace{-10px}
\caption{Syntax extension systems produce unreasonable program text.}
\vspace{-7px}
\label{fig:K-dialect}
\end{figure}

Consider, for example, the perspective of a programmer attempting to comprehend, i.e. reason about, the program text in Figure \ref{fig:K-dialect}, which is written in a dialect of OCaml's syntax called Reason \cite{reason-what}\footnote{This paper uses Reason 2.0 as 3.0 was released just before submission.} that has, hypothetically, been extended with some number of new literal forms by a grammar extension system like Camlp4 \cite{ocaml-manual} or Sugar*~\cite{erdweg2011sugarj,erdweg2013framework}. 
The literal form on Line 3 constructs an encoding of a query in the niche stack-based database query language K, using its intentionally terse notation \cite{Whitney:2001:LOR:376284.375783}. The problem is that a programmer unfamiliar with this notation cannot easily answer questions like the following:

\begin{enumerate}[nolistsep,leftmargin=10pt]
\item \textbf{(Responsibility)} Which syntax extension determined the expansion of the literal form on Line 3? Might activating a new extension generate a conflicting expansion for the same form (i.e. is there syntactic ambiguity)?
\item \textbf{(Segmentation)} Are the characters \li{x}, \li{R} and \li{2} on Line 3 parsed as spliced expressions (meaning that they appear directly in the underlying expansion), or are they parsed in some other way peculiar to this literal form (e.g. as operators in the K query language)?
\item \textbf{(Capture)} If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 2, or might it capture an unseen binding of the same identifier in the expansion of Line 3?
\item \textbf{(Context Dependence)} If \li{w}, on Lines 1-2, is renamed, could that possibly break Line 3 or change its meaning because the expansion assumes that \li{w} is in scope?% What other hidden assumptions might the desugaring be making?
\item \textbf{(Typing)} What type does \li{y} have?
\end{enumerate}

% In short, the problem is that library providers cannot reason modularly about syntactic determinism (because other extensions might define overlapping forms), and if the desugaring of the program text is held abstract, programmers can no longer reason about types, binding and segmentation (i.e. answer questions like those above.) This is burdensome at all scales, but particularly when programming in the large, where it is common to encounter a wide variety of libraries \cite{DBLP:conf/sac/LammelPS11}. 
Forcing the programmer to reason transparently to answer basic questions like these defeats the ultimate purpose of syntactic sugar: decreasing cognitive cost \cite{Green89}. Analagous problems do not arise when programming without syntax extensions in languages like ML and Scala --- programmers can reason lexically about where variables and other symbols are bound, and types mediate abstraction over function and module implementations \cite{B304}. % From this perspective, it is unsurprising that the use of Camlp4 has been deprecated in the OCaml system.

An approach that has made some headway against these problems is Omar et al.'s \emph{type-specific languages} (TSLs) \cite{TSLs}. With this approach, the language delegates control over the parsing and expansion of certain \emph{generalized literal forms} to a parse function associated with the type that the form is being checked against. Generalized literal forms syntactically subsume other literal forms because their bodies---the characters between the outer delimiters, here \li{[|} and \li{|]}---are left  unparsed by the fixed context-free syntax of the language (i.e. they are initially parsed like raw string literals.) With TSLs, Line 3 of our example would require a type annotation:
\begin{lstlisting}[numbers=none]
    let y : kquery = [|(!R)@&{&/x!/:2_!x}'!R}|]
\end{lstlisting}
The parse function associated with the \li{kquery} type would be invoked to parse and expand the literal body.

TSLs support reasoning abstractly about \textbf{Responsibility} and \textbf{Typing}, due to the type-driven dispatch mechanism, but the remaining reasoning principles were not fully and formally established in the prior work. Moreover, the prior work considered only a monomorphic, nominally-typed language with local type inference, leaving open a number of problems that came up as we considered integrating TSLs into Reason/OCaml. These included non-trivial problems related to support for structural types (e.g. arrow types), parameterized types, abstract types, modules, pattern notation and ML-style type inference.  Another problem was that parse functions and expansions were completely restricted from accessing libraries. Taken together, we found that TSLs as described were not suitable for integration into Reason/OCaml.

\paragraph{Contributions} The aim of this paper is to address all of these deficiencies, resulting in the first system for defining new literal notation that (1) provably maintains all five of the abstract reasoning principles just outlined; and (2) that is expressive enough for integration into Reason/OCaml and other full-scale statically typed languages. We evaluate the first claim by developing a series of formal calculi equipped with proofs of these reasoning principles, and the second claim with several non-trivial case studies that appear throughout the paper and involve all of the aforementioned language features. The empirical study in the prior work on TSLs \cite{TSLs} supports the claim that a macro-based mechanism for defining new literal notation is broadly useful, so this paper is targeted at language designers that seek a deeper understanding of the reasoning principles at play, particularly as they relate to features common to ML and other full-scale statically typed languages (e.g. Haskell, Scala, Rust, etc.)

\begin{figure}
\begin{lstlisting}
  let w = compute_w();
  let x = compute_x(w);
  let y = $kquery [|SURL(!R)@&{&/EURLxSURL!/:2_!EURLxSURL}'!R}EURL|];
\end{lstlisting}
\vspace{-10px}
\caption{TLMs make examples like the one from Figure \ref{fig:K-dialect} more reasonable.}
\vspace{-7px}
\label{fig:K-tsm-example}
\end{figure}

The most obvious point of departure between TSLs and \emph{typed literal macros} (TLMs), which are the topic of this paper, is that we separate out the mechanism for defining new literal notation from the type definition mechanism of the language. TLMs are instead explicitly named and applied. For example, Figure \ref{fig:K-tsm-example} applies a TLM named \li{$kquery} to a generalized literal form delimited by \li{[|} and \li{|]} to express the K query example just discussed. The applied TLM parses the literal body and generates an expansion statically. During the typing phase, a validation step ensures that programmers can reason abstractly about types, binding and segmentation. Taken together, the programmer can reason as follows:
\begin{enumerate}[nolistsep,leftmargin=10pt]
\item \textbf{(Responsibility)} The applied TLM, \li{$kquery}, is delegated responsibility for parsing the literal body. The context-free syntax of the language is fixed, so the TLM provider can modularly establish that the syntax that the TLM implements is unambiguous (even in situations where a spliced term itself applies a TLM, shown in a later example.)
\item \textbf{(Segmentation)} The intermediate output that the TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms from segments that are parsed in some other way by the TLM. The segmentation can be communicated by a code editor using secondary notation, e.g. colors in this document. So by examining Figure \ref{fig:K-tsm-example}, the programmer knows that the two instances of \li{x} on Line 3 are parsed as spliced expressions (because they are black), whereas the \li{R}'s must be parsed in some other way, e.g. as operators in the K language (because they are  green.)
\item \textbf{(Capture)} Splicing is capture-avoiding, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2. It cannot capture a coincidental binding of \li{x} in the expansion.
\item \textbf{(Context Dependence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, e.g., \li{w} is in scope. The client is free to rename \li{w}.
\item \textbf{(Typing)} A type annotation on the definition of \li{$kquery} (not shown) determines the type that the expansion must have. We will see an example of a TLM definition in the next section. Moreover, each spliced segment in the segmentation also comes with a type annotation. %This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired. %TLM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kquery}.
\end{enumerate}

@@

\begin{subfigure}[t]{\textwidth}
\begin{lstlisting}
  let heading first_name = $html [|SURL<h1>Hello, {[EURLfirst_nameSURL]}!</h1>EURL|]
  let body = $html [|SURL<body>{EURLheading "World!"SURL} ...</body>EURL|]
\end{lstlisting}
\vspace{-5px}
\caption{Two examples of the \li{$html} TLM being applied. Compare to Ur/Web's built in HTML literal notation from Figure \ref{fig:urweb}.  %In each case, literal body, between backticks, is initially left unparsed according to the language's context-free syntax. %The applied TLM determines a segmentation and expansion during the typed expansion phase, which generalizes the usual typing phase.
}
\label{fig:first-tsm-example}
\end{subfigure}
\vspace{-8px}
\caption{Case Study: HTML literals}
\vspace{-7px}
\end{figure*}

\vspace{-3px}
\paragraph{Outline} 
% \begin{itemize}
% \item 
We begin in Sec.~\ref{sec:setsms} by introducing TLMs with a detailed case study. 
% \item 
Sec.~\ref{sec:sptsms} then introduces {pattern TLMs}, i.e. TLMs that generate patterns, and describes the special hygiene conditions in pattern position. %The hygiene condition for pattern TLMs is interesting.
% \item 
Sec.~\ref{sec:ptsms} adds type functions and an ML-style module system and introduces \emph{parametric TLMs} (pTLMs), i.e. TLMs that can take type and module parameters, which solve several problems. 
Sec.~\ref{sec:setsms-formally} presents a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above. 
% \item 
Sec.~\ref{sec:existing-approaches} compares TLMs to related work using the rubric of five reasoning principles just discussed. 
% \item 
Sec.~\ref{sec:discussion} concludes with a summary of the specific contributions of this paper and a discussion.

@@

Due to the space limitations, certain technical details and extensions to the calculi presented in this paper, as well as detailed proofs, are given in the supplemental material. Also available is an additional series of case studies: notation for an encoding of regular expressions as an abstract data type, implemented using TLMs for quasiquotation and grammar-based parser generators. %are also available in the supplement. Sec.~\ref{sec:static-eval} considers the topic of term evaluation during the  expansion phase, i.e. \emph{static evaluation}, in  more detail. Sec \ref{sec:static-eval} also gives examples of TLMs that are useful for defining other TLMs, e.g. TLMs that implement parser generators and quasiquotation.

@@

\vspace{-6px}
\newcommand{\seTLMsSec}{Expression TLMs}
\section{\protect\seTLMsSec}
\label{sec:setsms}
% \vspace{-3px}

We start in this section by describing the TLM mechanism in detail by way of a substantial case study: literal notation, like that built in to Ur/Web (see Figure \ref{fig:urweb}) and also, presently, Reason, for expressions of the type  \li{html} in Figure \ref{fig:html-type-def}.

% \vspace{-8px}
\subsection{TLM Definition and Application}
\vspace{-2px}
The definition of the \li{$html} TLM is given in Figure \ref{fig:html-tlm-def}. TLM names are prefixed by \li{$} 
 to distinguish them from variables. Every TLM definition also includes a \emph{type annotation}, here \li{at html}, and a \emph{parse function} between \li{by static \{} and \li{\}}. % The TLM definition follows the same scoping rules as type definitions. We will describe how TLM definitions are packaged into libraries in Sec \ref{sec:static-eval}.

Using this TLM, we can express the Ur/Web example from Figure \ref{fig:urweb} as shown in Figure \ref{fig:first-tsm-example}. On both Lines 1 and 2, we apply 
\li{$html} 
 to a \emph{generalized literal form} delimited by \li{[|} and \li{|]}. Generalized literal forms, which first arose in the prior work on TSLs \cite{TSLs}, syntactically subsume other literal forms because the context-free syntax of the language only specifies the outer delimiters. In this paper, we will use  \li{[|} and \li{|]}, but \citet{TSLs} formally specified several other choices, including layout-sensitive delimitation. \emph{Literal bodies} are constrained only in that \li{[|} and \li{|]} must be balanced.  %In particular, there is no pre-specified syntactic to splice out sub-terms.%The initial context-free parser therefore parses generalized literal forms in much the same way as parsers typically parse ``raw'' string literals (i.e. string literals where escape sequences like \li{\\n} have not yet been processed.)

 The system delegates responsibility over the parsing and expansion of each literal body to the applied TLM's parse function during a semantic phase called \emph{expansion}, which starts just before and continues into the typing phase.
